---
title: "2021_0831_data_analysis"
author: "Raghav Jain"
date: "8/30/2021"
output: html_document
---

This code is associated with the manuscript: 

The below code was used to perform statistical analyses and create major figures for the paper. The data used for this analysis was pre-processed as described in the 'data_processing' code in the folder of that name. The example data being used is from plasma, but data for all tissue is also available in the 'raw data' folder.

Please send any questions to the developer of the script, Raghav Jain (jainragh52@gmail.com). Thank you to all the package developers and the broader R community for supporting open access and code availability.

# Load packages
```{r, message=FALSE, warning=FALSE, echo = F}
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(ggpubr)){install.packages("ggpubr")}
if(!require(reshape)){install.packages("reshape")}
if(!require(rstatix)){install.packages("rstatix")}
if(!require(ggrepel)){install.packages("ggrepel")}
if(!require(ggsci)){install.packages("ggsci")}
if(!require(scales)){install.packages("scales")}
if(!require(heatmaply)){install.packages("heatmaply")}
if(!require(factoextra)){install.packages("factoextra")}
if(!require(RColorBrewer)){install.packages("RColorBrewer")}
if(!require(outliers)){install.packages("outliers")}
#show_col(pal_npg("nrc")(10))
#get_palette(palette = "npg", 5)
```

# PCA
```{r}
mydata <- read_csv("raw_data/Lipid_data_all.csv")
mydata[mydata == 0] <- NA 

Y3 <- mydata[!mydata$Tissue == "Plasma",]

# Set the minimum number of samples that must have a value for a given lipid
missing.value = as.character()
for (i in 6:ncol(Y3)){
  if(table(is.na(Y3[,i]))[1] == 108)
    missing.value[i] = "NO"
  else(missing.value[i]="YES")
}

Y3_rm = data.frame(cbind(colnames(Y3), missing.value)) 
d <- Y3_rm[Y3_rm$missing.value=="YES", 1]

Y4 <- Y3[,!colnames(Y3) %in% d]

# RT
Y5 <- Y4[Y4$Temp == 'RT',]
P <- log2(Y5[,6:ncol(Y5)]) 
P <- P[,colSums(is.na(P))<nrow(P)]
res_pca <- prcomp(na.omit(P), center = T, scale. = TRUE)
fviz_eig(res_pca)

fviz_pca_ind(res_pca, title = 'RT',
             label = "none", habillage = Y5$Tissue, 
             addEllipses = T) + 
    theme(legend.position = "right", 
          text = element_text(size = 12, family = "Arial")) +
    theme_minimal()
```


# Heatmap comparing identified lipids
Heatmap Groups are:
-Acylcarnitines
-Cholesteryl Esters
-Fatty acids: FFA, FAHFA
-Glycerolipids: TAG, DAG
-Glycerophospholipids: PC,PE, PI, PS, PA, PG, (hemi-)BMP, CL, LPI, LPC, LPE, Ether-PC, Ether-PE
-Sphingolipids: SM, ceramides, hexosyl, Gangliosides, sulfatide

```{r}
Y <- read_csv("raw_data/identified_lipids.csv")
```

Visualize and output heatmap. Rename for different tissue. May need to change 'Blues' ramp for different tissue.
```{r}
Y0 <- Y[complete.cases(Y),]
Y1 <- aggregate(Y0[,5:ncol(Y0)], by=list("Lipid Class" = Y0$Class), FUN=sum)

# Reorder is based on PCA clustering
names(Y1)
Y1 <- Y1[c("Lipid Class", "Plasma", "BAT",
           "eWAT", "iWAT", "GSM", "Heart",
           "Liver", "Kidney", "Intestine", 
           "Lung")]
heatmaply(x = Y1[,2:ncol(Y1)], labRow = Y1$`Lipid Class`, color = Blues(180), 
          Colv = F, fontsize_row = 12, fontsize_col = 12, 
          font = list(family = "Arial", face = 2),
          cellnote = Y1[,2:ncol(Y1)], cellnote_size = 14, 
          cellnote_textposition = "middle center", 
          show_dendrogram = F, xlab = "Tissue", ylab = "Lipid Class", 
          main = "Tissue Lipidomics", 
          height = 650, width = 800, 
          file = NULL) 
```

# Segmented Bar Graphs
```{r,  message=FALSE}
Y <- read_csv("raw_data/identified_lipids.csv")
Y2 <- Y[,c(2,4:ncol(Y))]
Y2[is.na(Y2)] <- 0
Y3 <- reshape2::melt(Y2, id=c("Abbreviation","Bars"))

classes <- split(Y3, Y3$Bars)

# Tells where in the str 'classes' each class is
summary(classes)

# Create palette of 74 colors
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

my_color <- col_vector[!grepl('FF', col_vector)]                                  
# removes the yellow colors
```
Plot sphingolipid breakdown as an example.
```{r}
temp <- classes[[8]]
p0 <- ggplot(temp, aes(x = variable, y = value, fill = Abbreviation)) +
            geom_bar(stat = "identity", position = "stack")
  
p1 <- ggpar(p0, palette = my_color,legend = "right", legend.title = "Lipid Species", 
                  title = names(classes)[8], xlab = "Tissue", ylab = "Number of Species")
p1 + theme_bw()
```

# Volcano Plots
```{r}
Z <- read_csv("raw_data/2021_0917_plasma.csv")
Z[Z==0] <- NA
X = matrix(NA, ncol = 3, nrow = ncol(Z))
colnames(X) <- c("FC", "p", "sig")
rownames(X) <- colnames(Z)
  
p_value = as.numeric()

# Start at the lipid data for loop. Calculate fold change and p-values for comparison.
for (j in 6:ncol(Z)) {                               
    X[j,1] = round(colMeans(Z[Z$Temp == "Cold",j],na.rm = T)/colMeans(Z[Z$Temp == "RT",j], 
                                                                      na.rm = T), 4)
    
      if(table(is.na(Z[Z$Temp == "Cold",j]))[1] > 1 && table(is.na(Z[Z$Temp == "RT",j]))[1] > 1){
        X[j,2] = round(-log10(t.test(Z[Z$Temp == "Cold",j], Z[Z$Temp == "RT",j], 
                                     na.rm = T)$p.val), 4)
      }else(X[j,2] == 0.9)
      
      if(table(is.na(Z[Z$Temp == "Cold",j]))[1] > 1 && table(is.na(Z[Z$Temp == "RT",j]))[1] > 1){
        p_value[j] = t.test(Z[Z$Temp == "Cold",j], Z[Z$Temp == "RT",j], na.rm = T)$p.val
      }else(p_value[j] == 0.9)
}
  
# FDR correction applied  
op <- p.adjust(p_value[6:length(p_value)], method = 'fdr')
Y1 <- data.frame(X[6:nrow(X),])

# Set FDR significance cutoff
Y1$sig <- op < 0.300001
  
Y1$FC <- log2(Y1$FC)
  
Y1$diffexpressed <- "NO"
Y1$diffexpressed[Y1$FC > 0.0 & Y1$sig == T] <- "UP"
Y1$diffexpressed[Y1$FC < 0.0 & Y1$sig == T] <- "DOWN"
  
Y1$diffexpressed = factor(Y1$diffexpressed, levels = c("NO", "UP", "DOWN"))
  
Y1$delabel <- NA
Y1$delabel[Y1$diffexpressed != "NO"] <- rownames(Y1)[Y1$diffexpressed != "NO"]
  
toMatch <- c("ACar", "Cer")
  
Y1$delabel <- ifelse(grepl(paste(toMatch,collapse="|"), 
                             Y1$delabel),Y1$delabel,NA)
  
Y1$label <- NA
Y1$label[Y1$sig == T] <- "sig"
Y1$label[Y1$sig == F] <- "not"
  
ggplot(data=Y1, aes(x=FC, y=p, col=diffexpressed)) +
        geom_point() + 
        scale_x_continuous(limits = c(max(Y1$FC)*-1.1, max(Y1$FC)*1.1)) +
        scale_y_continuous(limits = c(0, max(Y1$p*1.1))) +
        theme_minimal() +
        geom_text_repel(aes(label=delabel), color = "black", min.segment.length = 0, 
                        size = 2.5, family = "Arial", max.overlaps = Inf) +
        scale_color_manual(values=c("grey70", "#E64B35FF", "#4DBBD5FF"), 
                           name = "Change in Cold",
                           breaks = c("NO", "UP", "DOWN"), 
                           labels = c("FDR > 0.30", "Increased", "Decreased")) +
        ggtitle("Plasma lipid changes") +
        ylab("-log10(p-value)") +
        xlab("log2(Fold change [cold/RT])") +
        theme(legend.position = "right", text = element_text(size = 12, 
                                                             family = "Arial"))
```

# Acylcarnitine bar graphs
Acar 16:0, 18:0, 18:1
```{r}
Z <- read_csv("raw_data/2021_0917_plasma.csv")
mdat <- Z
mdat[mdat == 0] <- NA
 
mdat1 <- cbind("Condition" = mdat$Temp, mdat[,grep("ACar", colnames(mdat))])
mdat1 <- mdat1[,colSums(is.na(mdat1)) < nrow(mdat1)]
mdat2 <- reshape2::melt(mdat1, id="Condition")
mdat2$Condition <- factor(mdat2$Condition, levels = c("RT", "Cold"))
mdat2$variable <- factor(mdat2$variable)
colnames(mdat2)[2] <- "Lipid"
mdat2$Lipid <- gsub("ACar", "", mdat2$Lipid)
mdat2 <- na.omit(mdat2)
  
p0 <- ggbarplot(mdat2, x = "Lipid", y = "value",
            add = c("mean_sd", "dotplot"), color = "Condition", fill = "Condition", 
            add.params = list(width = 0.35,
                              binwidth = .015*max(mdat2$value, na.rm = T)),
            alpha = 0.5, position = position_dodge(0.8), size = 0.15) +
            scale_y_continuous(limits = c(0.00, 1.15*max(mdat2$value, na.rm = T)),
                               expand = c(0,0))

p1 <- ggpar(p0, palette = "npg", legend = "right", legend.title = "Temperature", 
              title = "Plasma Acylcarnitines",  
              xlab = "Acylcarnitine", ylab = "pmol lipid/mg tissue") 
  
  stat.test <- mdat2 %>%
  group_by(Lipid) %>%
  na.omit() %>%
  t_test(value ~ Condition) %>%
  adjust_pvalue(method = "none") %>%
  add_significance("p")
  
  stat.test$p.adj.signif <- stat.test$p.signif
  stat.test <- stat.test %>%
                  na.omit() %>%
                  add_xy_position(fun = "mean_sd", "Lipid", dodge = 0.8) 

p1 + stat_pvalue_manual(
      stat.test, label = "p.signif", tip.length = 0.02, hide.ns = T)

```

# Correlation plot - ACar 18:0, 18:1, 16:0
```{r}
mydata <- read_csv("raw_data/Lipid_data_all.csv")

mdat1 <- cbind(mydata[mydata$Tissue == 'Liver',1:5],
                 mydata[mydata$Tissue == 'Liver',]$`ACar 18:0`)
colnames(mdat1)[6] <- "Liver_AC"
mdat1$Plasma_AC <-  mydata[mydata$Tissue == 'Plasma',]$`ACar 18:0`
mdat1$BAT_AC <- mydata[mydata$Tissue == 'BAT',]$`ACar 18:0`

mdat1$Temp <- factor(mdat1$Temp, levels = c("RT", "Cold"))
```

```{r}
# Liver
sp <- ggscatter(mdat1, x = "Liver_AC", y = "Plasma_AC", 
                color = "Temp", add = "reg.line",
                add.params = list(color = "black", fill = "lightgray"),
                conf.int = TRUE)

# Add correlation coefficient
sp + stat_cor(method = "pearson", 
                 label.x = min(mdat1$Liver_AC, na.rm = T), 
                 label.y = max(mdat1$Plasma_AC, na.rm = T), 
                 p.accuracy = 0.001, r.accuracy = 0.01) + 
        scale_colour_manual(name="Temperature",
                            labels=c("RT","Cold"),
                            values = c("#E64B35FF", "#4DBBD5FF")) +
        xlab("Liver (pmol/mg)") +
        ylab("Plasma (pmol/mL)") + ggtitle("Acylcarnitine 18:2") +
        theme_minimal() +
        theme(legend.position = "right", 
              text = element_text(size = 12, family = "Arial"),
              axis.line = element_line(colour = "black"))

# BAT
sp <- ggscatter(mdat1, x = "BAT_AC", y = "Plasma_AC", 
                color = "Temp", add = "reg.line",
                add.params = list(color = "black", fill = "lightgray"),
                conf.int = TRUE)

sp +  stat_cor(method = "pearson", 
                label.x = min(mdat1$BAT_AC, na.rm = T), 
                label.y = max(mdat1$Plasma_AC, na.rm = T), 
                p.accuracy = 0.001, r.accuracy = 0.01) +
        scale_colour_manual(name="Temperature", 
                      labels=c("RT","Cold"), 
                      values = c("#E64B35FF", "#4DBBD5FF")) +
        xlab("BAT (pmol/mg)") + 
        ylab("Plasma (pmol/mL)") + 
        ggtitle("Acylcarnitine 18:2") +
        theme_minimal() +
        theme(legend.position = "right", 
              text = element_text(size = 12, family = "Arial"),
              axis.line = element_line(colour = "black"))
```

# Acylcarnitine 18:1 Regressions
Data is log2 transformed, mean centered, and sd scaled.
Mouse 8 will continue to be excluded. Only lipids with values for all remaining samples regressed.

Below is example code for a regression predicting plasma ACar 18:1 with BAT ACar 18:1. 
```{r}
mdat <- read_csv("raw_data/Lipid_data_all.csv")
mdat[mdat==0] <- NA

Z <- mdat[mdat$Tissue == "Plasma",] 
P <- Z[c(1:7, 9:12),] # Exclude mouse 8 - plasma outlier

# specify y-variable to be predicted
y_var <- scale(log2(P$`ACar 18:1`), center = TRUE, scale = TRUE)
Cond1 <- P$Temp
Z1 <- mdat[mdat$Tissue == "BAT",] 
Z1[Z1==0] <- NA
B <- Z1[c(1:7, 9:12),]
x_var <- scale(log2(B$`ACar 18:1`), center = TRUE, scale = TRUE)


my_lm <- lm(y_var ~ Cond1 + x_var)

lipid <- "ACar 18:1"
tissue <- "BAT"
        
y_int <- summary(my_lm)$coefficients[1,1]
beta_RT <- summary(my_lm)$coefficients[2,1]
beta_lipid <- summary(my_lm)$coefficients[3,1]
        
p_y_int <- summary(my_lm)$coefficients[1,4]
p_RT <- summary(my_lm)$coefficients[2,4]
p_lipid <- summary(my_lm)$coefficients[3,4]
    
adj_r_sq <- summary(my_lm)$adj.r.sq
    
final_data <- cbind(lipid, tissue, y_int, beta_RT,
                          beta_lipid, p_y_int, p_RT, 
                          p_lipid, adj_r_sq)

final_data <- as_tibble(final_data)
```

